\thetitle

### 2021-03-04 writing/challenges.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a set of
components. The Erasure semantics does not blame components {\it per se\/}, but
it comes with an exception location and a stack trace, which implicitly suggests
parties to blame.  Each strategy affects the behavior of the rational programmer
differently (and real ones, too).

One way to reconcile these differences is to {\em equip the rational
programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Intuitively, with the common vocabulary of modes, different
blame strategies become different ways the rational programmer  interprets
blame.  For instance, one Transient mode may pick the oldest element of a
blame set as the next component to add types because it corresponds to the
earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the set as a stack and add
types to its newest element.  If both modes are equally successful in
locating an impedance mismatch, measuring the rational programmer's effort with each mode
may answer which is the most effective.

However, attributing the success of the rational programmer to this or
that blame mechanism demands a careful look at the interplay between blame
and the run-time checks of each gradual type system. When a check fails
the Natural and Transient semantics assign blame instead of using the
information in the exception from the run-time check. Nevertheless, the
exception information may be equally or even more useful to the rational
programmer than blame. If this is the case, then blame may not play a critical
role for the rational programmer. Indeed, precisely because they do not
account for such confounding factors, \citet{lksfd-popl-2020} cannot draw any
conclusions about blame specifically despite advertisements for the
opposite. Their experiment can conclude that so-called blame-shifting
works but not that this is because of blame.

Modes offer a uniform way to compare the different semantics and  isolate
blame from the effect of the semantics' run-time checks. Specifically, the Natural and
Transient rational programmer come with a blame mode and an exception
mode. If the former succeeds debugging a program while the latter
fails, then it is safe to conclude that blame is indeed necessary for the
rational programmer. Put differently, the exception mode serves as the
{\em baseline\/} for blame's value within a given semantics; if the
programmer in this mode performs as well or better than the  blame one, a blame assignment
mechanism might be useless.

@@

An alternative is to {\em generate a large and representative corpus of type
mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
analysis traditionally aims to inject bugs that challenge test suites, and it
discards those that yield ill-typed mutants as \emph{incompetent}. Indeed,
mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
precisely those that are needed for evaluating blame assignment strategies.

@@

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutated component.  A
debugging scenario results from removing the types from the mutated component. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About four fifths of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

>>>>

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame components {\it per se\/}, but
it comes with an exception location and a stack trace, which implicitly suggests
parties to blame.  Each strategy affects the behavior of the rational programmer
differently (and real ones, too).

One way to reconcile these differences is to {\em equip the rational
programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Intuitively, with the common vocabulary of modes, different
blame strategies become different ways the rational programmer  interprets
blame.  For instance, one Transient mode may pick the oldest element of a
blame sequence as the next component to add types because it corresponds to the
earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the sequence as a stack and add
types to its newest element.  If both modes are equally successful in
locating an impedance mismatch, measuring the rational programmer's effort with each mode
may answer which is the most effective.

However, attributing the success of the rational programmer to this or
that blame mechanism demands a careful look at the interplay between blame
and the run-time checks of each gradual type system. When a check fails
the Natural and Transient semantics assign blame instead of using the
information in the exception from the run-time check. Nevertheless, the
exception information may be as useful to the rational
programmer as a blame assignment. If this is the case, then blame may not play a critical
role for the rational programmer. Indeed, precisely because they do not
account for such confounding factors, \citet{lksfd-popl-2020} cannot draw any
conclusions about blame specifically, despite advertisements for the
opposite. Their experiment may conclude only that so-called blame-shifting
works, but they cannot attribute this conclusion to blame.

Modes offer a uniform way to compare the different semantics and  isolate
blame from the effect of the semantics' run-time checks. Specifically, the Natural and
Transient rational programmer come with a blame mode and an exception
mode. If the former succeeds in debugging a program while the latter
fails, then it is safe to conclude that blame is indeed necessary for the
rational programmer. Put differently, the exception mode serves as the
{\em baseline\/} for blame's value within a given semantics; if the
programmer in this mode performs as well or better than the  blame one, a blame assignment
mechanism might be useless.

@@

An alternative is to {\em generate a large and representative corpus of type
mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
analysis traditionally aims to inject bugs that challenge test suites, and it
discards those that yield ill-typed mutants as \emph{incompetent}. Indeed,
mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
precisely those mutators that are needed for evaluating blame assignment strategies.

@@

An experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutated component.  A
debugging scenario results from removing the types from the mutated
component. For the design of such mutators, the authors relied on their own
extensive programming experience though not without discovering a major pitfall:
some of their original mutators systematically produced programs that
immediately revealed the source of the impedance mismatch.  All of the remaining
ones yield {\em interesting debugging scenarios\/} (see
sec.~\ref{sub:mutate-interesting}).

### 2021-03-04 writing/discussion.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

The results in section~\ref{sec:results} show that Transient produces empty
blame sets for 967 scenarios. In theory, empty set must not occur. An
empty blame set means a lack of boundary crossings for the witness value.
By implication, a typed module is blaming itself---something that can
happen only if the type checker (or system) is unsound.

@@

{\bf Note} The scenarios where Transient returns empty blame 
suggest ways to improve Transient's blame map as follows:
\begin{itemize}

>>>>

The results in section~\ref{sec:results} show that Transient produces empty
blame sequences for 967 scenarios. In theory, empty set must not occur. An
empty blame sequence means a lack of boundary crossings for the witness value.
By implication, a typed module is blaming itself---something that can
happen only if the type checker (or system) is unsound.

@@

{\bf Note} The scenarios where Transient returns no blame 
suggest ways to improve its blame map:
\begin{itemize}

### 2021-03-04 writing/introduction.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic
systems (Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies (see
sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches:
disagreements between the type ascriptions of a software system and its
behavior.

@@

This contrast between theory and applications of gradual typing raises the question of
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language,
 especially for the benefit of the working programmer.
\end{quote}
Given the long-standing academic interest in blame and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not discuss any possible answers.
Instead, when language designers make relevant decisions, they seem to answer it
one way or another without any scientific justification. Then again, the
community has thus far failed to offer a method for evaluating blame assignment.

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be implemented and
tested at scale on real programs.  Turning this idea into a scientific
experiment requires overcoming major challenges: injecting representative
impedance mismatches; putting the various kinds of error information to
comparable use; and sampling the huge space of possibilities (see
sec.~\ref{sec:challenges} and, for details, secs.~\ref{sec:rational} through~\ref{sec:sample}).

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct blame assignment
strategies in approximately 72,200 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
First, a good blame assignment strategy helps with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Second, Natural's wrapper-based blame tracking is
more useful than Transient's ``collective blame'' tracking algorithm,
which in turn is superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. The
first problem concerns the difference between the methods; neither is {\it
Natural\/} vastly more useful than {\it Transient\/} nor are the two
blame assignment methods always superior to {\it Erasure\/}.  The second
one is that the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need
additional experiments to collect more data.

>>>>

%% ------------------------------------------------------------------------------

Theoreticians of gradual typing have focused on blame theorems from the
very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic
systems (Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017,
vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
come with sophisticated checking and blame assignment strategies (see
sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
blame can help practicing programmers find impedance mismatches, that is,
disagreements between the type ascriptions of a software system and its
behavior.

@@

This contrast between theory and applications of gradual typing raises the question of
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language,
 especially for the benefit of the working programmer.
\end{quote}
Given the long-standing academic interest in blame and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not discuss any possible answers.
Instead, when language designers make decisions concerning this aspect, they seem
to go one way or another without any scientific justification. Then again, the
community has thus far failed to offer a method for evaluating blame assignment.

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be implemented and
tested, at scale, on real programs.  Turning this idea into a scientific
experiment requires overcoming major challenges: injecting representative
impedance mismatches; putting the various kinds of error information to
comparable use; and sampling the huge space of possibilities (see
sec.~\ref{sec:challenges} and, for details, secs.~\ref{sec:rational} through~\ref{sec:sample}).

The paper's {\it second contribution\/} is a set of {\em results from
applying the evaluation method to three distinct blame assignment
strategies in approximately 72,000 different scenarios\/} (see
sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
its blame assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
the approach of industrial systems, which forgo blame in favor of error
messages from the safety checks of the underlying language. The results
(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
principle they validate the conjectures behind the work of theoreticians.
First, a good blame assignment strategy helps with the search for
impedance mismatches between the specified types and the behavior of
untyped components.  Second, Natural's wrapper-based blame tracking is
more useful than Transient's ``collective blame'' tracking algorithm,
which in turn is superior to Erasure. {\em But\/}, the application of the
method also indicates problems with the expectations of theoreticians. The
first problem concerns the difference between the methods; neither is {\it
Natural\/} vastly more useful than {\it Transient\/} nor are the two
blame assignment methods highly superior to {\it Erasure\/}.  The second
one is that the cost of Transient's blame can be huge.  In turn, these
problems suggest that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need
additional experiments to collect more data.

### 2021-03-04 writing/landscape.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
among others, a function {\tt pack}. As the documentation says, this function
consumes JSON data and packages it in a dictionary. By mistake, though, the
function returns a hash table instead of the association list its documentation
promises.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function with run-time checks.
In essence, these checks protect typed modules when they call an untyped function
because it may not produce the promised kind of value, as is the case here with
{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps each exported
function of a typed module (such as {\tt crypto-pack}) in a proxy that checks
its arguments.  These checks protect functions exported from typed modules against
applications in untyped code that supply the wrong kind of arguments.

As this analysis implies, if a return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
an argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails and Natural blames
\texttt{pack-lib}.

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the contents of a
complex value are checked on a piecemeal basis.

@@

In order to locate the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed and untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt typed-pack-lib} and
from {\tt pack-lib} to {\tt crypto-pack-lib} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

>>>>

\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
among others, a function {\tt pack}. As the documentation says, this function
consumes JSON data and packages it in an association list (like a dictionary). By mistake, though, the
function returns a hash table instead of the association list its documentation
promises.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function arguments with run-time checks.
Analogously the run-time system wraps each exported function of a typed module
(such as {\tt crypto-pack}) in a proxy that checks its arguments.  These checks
protect functions exported from typed modules against applications in untyped
code that supply the wrong kind of arguments.

As this analysis implies, if a return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
an argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails and Natural blames
\texttt{pack-lib} for both {\tt define}s in {\tt client}.

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the contents of a
complex value is checked on a piecemeal basis.

@@

In order to locate the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed and untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt typed-pack-lib} and
from {\tt pack-lib} to {\tt crypto-pack-lib} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame sequence that excludes responsible parties and includes modules
irrelevant to the failing check.

### 2021-03-04 writing/mutate-benchmarks.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

\item have a decent number of modules and a variety of module dependency graphs
 (since mixing of typed and untyped code in Typed Racket takes place at the
 module level);

\item attempt to take  advantage of the sophisticated typing features of
 a gradually typed language such as Typed Racket but;

\item  fail and contain type-level mistakes. 
\end{enumerate}
\noindent
No such benchmark suite exists.  Worse, no benchmark suite of
gradually typed programs with type-level mistakes exists at
all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
Python programs to evaluate their technique of fixing mistakes in type
annotations. Their collection does not come with type-level mistakes
in the code itself.}

Fortunately, \citet{gtnffvf-jfp-2019} have curated a collection of Typed Racket
programs for systematically measuring the implementation's performance. The
benchmark suite consists of fully typed, correct programs, written by a number
of different authors, who have maintained and evolved these programs over
several years. The programs range widely in size, complexity, purpose, origin
and in programming style. The latter is mostly a reference to their reliance on
many Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for
mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
numeric tower~\cite{stathff-padl-12}.

Hence, even if they do not come with type-level errors, they offer a
diverse linguistic substrate suitable for the careful introduction of
type-level errors via mutations. Further filtering the benchmark suite to
keep the benchmarks with the most sizeable and complex dependency graphs
yields ten suitable programs while preserving the suite's linguistic
diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
benchmark programs and describes their basic characteristics.

>>>>

\item have a decent number of modules and a variety of module dependency graphs
 (because mixing of typed and untyped code in Typed Racket takes place at the
 module level);

\item take advantage of the variety of typing features of a gradually typed
 language; but

\item fail and contain type-level mistakes. 
\end{enumerate}
No such benchmark suite exists.  Worse, no benchmark suite of
gradually typed programs with type-level mistakes exists at
all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
Python programs to evaluate their technique of fixing mistakes in type
annotations. Their collection does not come with type-level mistakes
in the code itself.}

What exists is \citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs
for systematically measuring the implementation's performance. The benchmark
suite consists of fully typed, correct programs, written by a number of
different authors, who have maintained and evolved these programs over several
years. The programs range widely in size, complexity, purpose, origin and in
programming style. The latter is mostly a reference to their reliance on many
Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's numeric
tower~\cite{stathff-padl-12}.

Hence, even if they do not come with type-level errors, they offer a diverse
starter suite, suitable for the careful introduction of type-level errors via
mutations. Further filtering the benchmark suite to keep the benchmarks with the
most sizeable and complex dependency graphs yields ten suitable programs while
preserving the suite's linguistic diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
benchmark programs and describes their basic characteristics.

### 2021-03-04 writing/mutate-interesting.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

{\bf Note} While this choice favors Erasure over Transient and Natural and, for
the same reason, Transient over Natural, some form of bias towards one or the
other semantics is unavoidable. Tipping the scales in favor of the theoretically
weakest semantics yields the most stable results. See
section~\ref{sec:discussion} for some further discussion of this choice.

@@

\caption{The mutators produce Interesting debugging scenarios across all benchmarks.} 
 \label{fig:mutant-breakdown}
\end{figure*}

@@

The definition of interesting mutants is more than a filter for
uninteresting mutants. It is also the guide for evaluating and refining
mutators until they collectively produce a sizable and diverse population of
interesting mutants. For example, one early mutator idea casts the
tests of conditional expressions to the {\tt Any} type to
suppress occurrence typing. However, an examination of the mutants it
produces through the lens of the definition of interesting mutants
reveals that none is interesting because they never raise run-time errors.
Indeed, the interesting standard guided countless iterations of adding,
removing, and refining the mutators in figure~\ref{table:mutation-ops}
to maximize interesting mutants.

>>>>

{\bf Note} While this choice favors Erasure over Transient and Natural and, for
the same reason, Transient over Natural, some form of bias towards one or the
other semantics is unavoidable. Tipping the scales in favor of the theoretically
weakest semantics yields the most stable results. 
Section~\ref{sec:discussion} includes some further discussion of this choice.

@@

\caption{The generated debugging scenarios across all benchmarks} 
 \label{fig:mutant-breakdown}
\end{figure*}

@@

The definition of interesting mutants is more than a filter for uninteresting
mutants. It is also the guide for evaluating and refining mutators until they
collectively produce a sizable and diverse population of interesting
mutants. Consider a candidate for the set of mutators that casts the tests of conditional
expressions to the {\tt Any} type, which would suppress occurrence typing. An
examination of its mutants through the lens of ``interesting'' reveals that none
of them is interesting, because they never raise run-time errors. Hence Indeed,
the interesting standard guided countless iterations of adding and removing
mutators and of refining the chosen mutators in figure~\ref{table:mutation-ops}.

### 2021-03-04 writing/mutate-mutators.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
constraints. As the explanation indicates, some specialize or
generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
the vast literature on mutation testing~\cite{jia2011analysis}.\footnote{Many
of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints and had to
be eliminated or significantly adapted. For example, the relational mutator does not lead to type-level
errors. Similarly, the arithmetic mutator does not reliably lead to type
errors; it had to be adapted to make only operator swaps that can affect the
type of the result, e.g., changing `*` to `/`.}  Only two are directly
inherited; many mutators are brand new. For the latter the authors relied on
their decades-long experience of making type-level mistakes in Typed Racket,
some of which take non-trivial effort to debug.

>>>>

%% MF: I don't see what this contributes:
% \footnote{Many of~\citet{lksfd-popl-2020} mutators do not satisfy the
% constraints and had to be eliminated or significantly adapted. For example, the
% relational mutator does not lead to type-level errors. Similarly, the
% arithmetic mutator does not reliably lead to type errors; it had to be adapted
% to make only operator swaps that can affect the type of the result, e.g.,
% changing `*` to `/`.}

Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
constraints. As the last column indicates, some specialize or
generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
the vast literature on mutation testing~\cite{jia2011analysis}.  Only two are
directly inherited; many mutators are brand new. For the latter, the authors
relied on their decades-long experience of making type-level mistakes in Typed
Racket, some of which take non-trivial effort to debug.

### 2021-03-04 writing/mutate.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

Putting the rational programmer to work means generating many mutants and turning
those into debugging scenarios. The process must start with a suitable
collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
Since existing mutators do not generate useful mutants, the first step is to
develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
suitability for this application (see sec.~\ref{sub:mutate-interesting}).

>>>>

Putting the rational programmer to work means generating many mutants and turning
those into debugging scenarios. The process must start with a suitable
collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
Since existing mutators do not generate useful mutants, the first step is to
develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
suitability on the benchmarks (see sec.~\ref{sub:mutate-interesting}).

### 2021-03-04 writing/paper.tex
re-established the old organization
editiroal voice thru 6
pass over challenges section
adjusting title for section 7
uniform editorial voice, mistakes, typos, size of text in figures, captions

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems
and prove blame theorems for gradually typed programming languages.
Practical implementations of gradual typing almost completely ignore
the idea of blame assignment.  This contrast raises the question
whether blame assignment provides any value to the working programmer
and poses the challenge of how to evaluate the effectiveness of blame
assignment strategies. This paper contributes (1) the first evaluation
method for blame assignment strategies and (2) the results from applying
it to three different gradual type systems. These results confirm that blame assignment
is helpful to a rationally acting programmer.
\end{abstract}

@@

\input{landscape-program}
\sec{introduction}{Does Blame Matter}
\sec{landscape}   {How to Think About Three Blame Systems}
\sec{why-rational}{What Is a Rational Programmer}
\sec{challenges}  {Why It Is Hard to Implement the Rational Programmer}
\sec{mutate}      {How to Make Lots of Mistakes}
\sec{rational}    {How to Make Comparable Rational Programmers}
\sec{sample}      {How to Make the Experimental Space}
\sec{results}     {What We Discovered}
\sec{discussion}  {What We Learned}
\sec{related}     {Who has Worked on this Problem Before}
\sec{conclusion}  {What to Do Next}
%% -----------------------------------------------------------------------------

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems
and prove blame theorems for gradually typed programming languages.
Practical implementations of gradual typing almost completely ignore
the idea of blame assignment.  This contrast raises the question
whether blame assignment provides any value to the working programmer
and poses the challenge of how to evaluate the effectiveness of blame
assignment strategies. This paper contributes (1) the first evaluation
method for blame assignment strategies and (2) the results from applying
it to three different gradually typed languages. These results confirm that blame assignment
is helpful to a rationally acting programmer.
\end{abstract}

@@

\sec{introduction}{Does Blame Matter}
\input{landscape-program}
\sec{landscape}   {How to Think About Three Blame Systems}
\sec{why-rational}{What Is a Rational Programmer}
\sec{challenges}  {Why It Is Hard to Evaluate Blame}
\sec{rational}    {How to Make Comparable Rational Programmers}
\sec{mutate}      {How to Make Lots of Mistakes}
\sec{sample}      {How to Make the Experimental Space Tractable}
\sec{results}     {What Are the Outcomes of the Experiment}
\sec{discussion}  {What Can Programmers Learn}
\sec{related}     {Who has Worked on this Problem Before}
\sec{conclusion}  {What to Do Next}
%% -----------------------------------------------------------------------------

### 2021-03-04 writing/rational-effort.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

In addition to success and failure information for the various trails, the
experiment records the number of modules a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as an additional metric to compare modes.

@@

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
module to type randomly.  Since each mutant has a finite number of
modules, random mode trails are always successful. However, the random mode's effort
distribution will be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

>>>>

In addition to success and failure information for the various trails, the
experimental test bed can record the number of modules a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as an additional metric to compare modes.

@@

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
module to type randomly.  Since each mutant has a finite number of
modules, random mode trails are always successful. However, the random mode's effort
distribution should be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

### 2021-03-04 writing/rational-experiment.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and the experiment
uses the Natural blame and exception trails to answer it, so $Q_1$ shows up in the cells for Natural blame and Natural exceptions.

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

\begin{figure}[bh]
  \centering
  \includegraphics[scale=0.36]{./Images/process}
  \caption{The experimental process for one mode of the rational programmer}
  \label{fig:process}
\end{figure}

>>>>

\begin{figure}[ht] \footnotesize
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \label{fig:experiment-outline}
\end{figure}

The above table summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example,
experimental question $Q_1$ asks whether blame is valuable for Natural and the
experiment uses the Natural blame and exception trails to answer it, so $Q_1$
shows up in the cells for Natural blame and Natural exceptions.

@@

\begin{figure}[bh] \footnotesize
  \centering
  \includegraphics[scale=0.36]{./Images/process}
  \caption{The experimental process for one mode of the rational programmer}
  \label{fig:process}
\end{figure}
